Alone Together: Compositional Reasoning and Inference for Weak Isolation

### 2017-11-06 implementation.tex
Typos etc.

\textbf{Annotation Burden} \thetool significantly reduces the annotation
burden in verifying a weakly isolated transactions by eliminating the
need to annotate intermediate assertions and loop invariants.
Guarantees ($G$) and global invariants ($I$), however, still need to be
provided. Alternatively, a weakly isolated transaction $T$ can be
verified against a generic serializability condition,  eliminating
the need for any annotation. In this mode, \thetool first infers the
transformer $\F_{SER}$ of $T$ without considering any interference,
which then becomes its guarantee ($G$). Doing likewise for every
transaction results in a rely relation ($R$) that includes $\F_{SER}$
of every transaction. Verification now proceeds by taking interference
into account, and verifying that each transaction still yields the
same $F$ as its $F_{SER}$. The result of this verification is an
assignment of (possibly weak) isolation levels to transactions which
nonetheless guarantees behavior equivalent to a  serializable execution.

>>>>

\textbf{Annotation Burden} \thetool significantly reduces the
annotation burden in verifying a weakly isolated transactions by
eliminating the need to annotate intermediate assertions and loop
invariants.  Guarantees ($G$) and global invariants ($I$), however,
still need to be provided. Alternatively, a weakly isolated
transaction $T$ can be verified against a generic serializability
condition,  eliminating the need for guarantee annotations. In this
mode, \thetool first infers the transformer $\F_{SER}$ of $T$ without
considering any interference, which then becomes its guarantee ($G$).
Doing likewise for every transaction results in a rely relation ($R$)
that includes $\F_{SER}$ of every transaction. Verification now
proceeds by taking interference into account, and verifying that each
transaction still yields the same $F$ as its $F_{SER}$. The result of
this verification is an assignment of (possibly weak) isolation levels
to transactions which nonetheless guarantees behavior equivalent to a
serializable execution.

### 2017-11-06 inference.tex
Typos etc.

  \C{del} field set to true in the local store.}. Note that $\F$
actually captures the \emph{effect} part of the state transformer,
but we nonetheless refer to it as a transformer. Since the execution
is subject to isolation-constrained interference, the inference
judgment depends on the isolation-constrained rely relation $\R$,
which is used to enforce the stability of the state transformer $\F$.
Recall that $\R$ is a tri-state rely relation over $\stl$, $\stg$ and
$\stg'$, that admits an interference from $\stg$ and $\stg'$ depending
on the local database state $\stl$. Thus, the stability of the state
transformer $\F$ of $c$ with respect to $\R$ needs to take into
account the (possible) prior state of the local database $\stl$, which
depends on the context (sequence of previous commands) of $c$, and
computed by the corresponding state transformer $\Fx$. Thus, the
semantics of the state transformer can be understood in terms of the
RG judgment as following (formalized as
Theorem~\ref{thm:inference-sound} in Sec.~\ref{sec:inference-sound}):

@@

The state transformer rules, like the earlier RG rules, closely follow
the corresponding reduction rules in Fig.~\ref{fig:txnimp}, except
that their language of expression is $\SL$. For instance, while the
reduction rule for \C{UPDATE} declaratively specifies the set of
updated records, the state transformer rule uses $\SL$'s bind
operation to \emph{compute} the set. Other SQL rules do likewise. The
rules for \C{LET} binders, conditionals, and sequences compose the
effects inferred for their subcommands. Thus, the effect of a sequence
of commands $c_1;c_2$ is the union of effects $\F_1$ and $\F_2$ of
$c_1$ and $c_2$, respectively, except that $\F_2$ is computed in a
context that includes $\F_1$ (we write $\F_1 \cup \F_2$ as a shorthand
for $\lambda(\stg).~\F_1(\stg) \cup \F_2(\stg)$). The inference rule
for \C{FOREACH} takes advantage of the $\SL$'s bind operator to lift
the effect inferred for the loop body to the level of the loop. Since
records added to $\stl$ in each iteration of \C{FOREACH} are
independent of the previous iteration (recall that we make a local
context independence assumption about database programs; Sec.
~\ref{sec:opsem}), sequential composition of the effects of different
iterations is the same as their parallel composition. Since the loop
body is executed once per each $z\in x$, the effect of the the loop is
a union of effects ($\F$) for all $z\in x$, all applied to the same
state ($\stl$ and $\stg$).  That is, $\F_{loop}(\stg) = \bigcup_{z\in
x}\F_{body}(\stg)$. From the definition of the set monad's bind
operator, $\F_{loop}(\stg) = x \bind (\lambda z.~F_{body}(\stg))$,
which mirrors the definition of the rule.
%This development mirrors ideas found in type checkers for certain
%dependent type systems~\cite{KJ14}.

@@

      (\lambda\C{item\_req}.~ \F_U(\stl,\stg) \cup \F_I(\stl,\stg))

@@

    \F_U & = & \lambda(\stl,\stg).~ \stg \bind(\lambda s. 

@@

         &   & \hspace*{0.9in}}

@@

         &   & \hspace*{0.9in}} {\emptyset}) \\
    \F_I & = & \lambda(\stl,\stg).~ \{\langle\C{ol\_o\_id}=\C{k2},\,

@@

For any $\stl$ and $\stg$, $\F_U(\stl,\stg)$ and $\F_I(\stl,\stg)$ are
expressions in $\SL$, so can be translated to FOL by the encoding
algorithm in Fig.~\ref{fig:logic}. Since the iteration variable
\C{item\_req} occurs free in these expressions, the appropriate
$\mssemof{\C{item\_req}}{\F_U(\stl,\stg)}$ and
$\mssemof{\C{item\_req}}{\F_I(\stl,\stg)}$, which results in

@@

Next, in Fig.~\ref{fig:logic}, observe that the (ordered) set $\nubar$
is extended only in the encoding rule for $\bind$. Since an occurrence
of $\bind$ adds a quantifier to $|\nubar|$, if $s$ is a bind
expression nested inside a top-level bind expression (like
$\F_U(\stl,\stg)$ from the above example), then the prenex signature
of $\phi_s$ is $\forall^2\exists$.  Furthermore, if the subexpressions
of $s$ are neither {bind} nor $\existsl$ expressions, then none of the
$\phi_i$'s are quantified, and the prenex signature of $\phi = \bigwedge_i\phi_i \wedge \phi_s$
remains $\forall^2\exists$. A similar
observation holds when $s$ is an $\existsl$ expression nested inside a
top-level {bind} expression.  Since $\existsl$ is generated as a
result of stabilizing a SQL command transformer, which is always a
non-nested bind expression, the subexpression ($s'$) of $\existsl$ is
a non-nested bind expression. $s'$ is however nested inside a
top-level bind expression, hence its prenex signature is
$\forall^2\exists$.  Since $\existsl$ doesn't extend $\nubar$, the
prenex signature of $s$ remains $\forall^2\exists$. When $s$ is an
expression other than $\bind$ or $\existsl$, then $\phi_s$ is not a
quantified formula, and its prenex signature is trivially subsumed by
$\forall^2\exists$. Thus, for the subset of $\SL$, where bind
expressions are restricted to one level of nesting, the FOL formulas
generated by the encoding have the prenex signature as
$\forall^2\exists$.

>>>>

  \C{del} field set to true in the local store.}. Note that the
function $\F$ we call state transformer here is actually the
\emph{effect} part of the state transformer introduced in
Sec.~\ref{sec:motivation}, which is a function $\T$ of form
$\lambda(\stl,\stg).~\stl \cup \F(\stg)$. Nonetheless, we keep
referring to $\F$ as state transformer. Since the execution is subject
to isolation-constrained interference, the inference judgment depends
on the isolation-constrained rely relation $\R$, which is used to
enforce the stability of the state transformer $\F$.  Recall that $\R$
is a tri-state rely relation over $\stl$, $\stg$ and $\stg'$, that
admits an interference from $\stg$ and $\stg'$ depending on the local
database state $\stl$. Thus, the stability of the state transformer
$\F$ of $c$ with respect to $\R$ needs to take into account the
(possible) prior state of the local database $\stl$, which depends on
the context (sequence of previous commands) of $c$, and computed by
the corresponding state transformer $\Fx$. Thus, the semantics of the
state transformer can be understood in terms of the RG judgment as
following (formalized as Theorem~\ref{thm:inference-sound} in
Sec.~\ref{sec:inference-sound}):

@@

The state transformer rules, like the earlier RG rules, closely follow
the corresponding reduction rules in Fig.~\ref{fig:txnimp}, except
that their language of expression is $\SL$. For instance, while the
reduction rule for \C{UPDATE} declaratively specifies the set of
updated records, the state transformer rule uses $\SL$'s bind
operation to \emph{compute} the set. Other SQL rules do likewise. The
rules for \C{LET} binders, conditionals, and sequences compose the
effects inferred for their subcommands. Thus, the effect of a sequence
of commands $c_1;c_2$ is the union of effects $\F_1$ and $\F_2$ of
$c_1$ and $c_2$, respectively, except that $\F_2$ is computed in a
context that includes $\F_1$ (we write $\F_1 \cup \F_2$ as a shorthand
for $\lambda(\stg).~\F_1(\stg) \cup \F_2(\stg)$). The inference rule
for \C{FOREACH} takes advantage of the $\SL$'s bind operator to lift
the effect inferred for the loop body to the level of the loop. Since
records added to $\stl$ in each iteration of \C{FOREACH} are
independent of the previous iteration (recall that we make a local
context independence assumption about database programs; Sec.
~\ref{sec:opsem}), sequential composition of the effects of different
iterations is the same as their parallel composition. Since the loop
body is executed once per each $z\in x$, the effect of the the loop is
a union of effects ($\F$) for all $z\in x$, all applied to the same
state ($\stg$).  That is, $\F_{loop}(\stg) = \bigcup_{z\in
x}\F_{body}(\stg)$. From the definition of the set monad's bind
operator, $\F_{loop}(\stg) = x \bind (\lambda z.~F_{body}(\stg))$,
which mirrors the definition of the rule.
%This development mirrors ideas found in type checkers for certain
%dependent type systems~\cite{KJ14}.

@@

      (\lambda\C{item\_req}.~ \F_U(\stg) \cup \F_I(\stg))

@@

    \F_U & = & \lambda(\stg).~ \stg \bind(\lambda s. 

@@

         &   & \hspace*{0.8in}}

@@

         &   & \hspace*{0.8in}} {\emptyset}) \\
    \F_I & = & \lambda(\stg).~ \{\langle\C{ol\_o\_id}=\C{k2},\,

@@

For any $\stg$, $\F_U(\stg)$ and $\F_I(\stg)$ are expressions in
$\SL$, so can be translated to FOL by the encoding algorithm in
Fig.~\ref{fig:logic}. Since the iteration variable \C{item\_req}
occurs free in these expressions, the appropriate application of the
encoding algorithm is $\mssemof{\C{item\_req}}{\F_U(\stg)}$ and
$\mssemof{\C{item\_req}}{\F_I(\stg)}$, which results in

@@

Next, in Fig.~\ref{fig:logic}, observe that the (ordered) set $\nubar$
is extended only in the encoding rule for $\bind$. Since an occurrence
of $\bind$ adds a quantifier to $|\nubar|$, if $s$ is a bind
expression nested inside a top-level bind expression (like
$\F_U(\stg)$ from the above example), then the prenex signature
of $\phi_s$ is $\forall^2\exists$.  Furthermore, if the subexpressions
of $s$ are neither {bind} nor $\existsl$ expressions, then none of the
$\phi_i$'s are quantified, and the prenex signature of $\phi = \bigwedge_i\phi_i \wedge \phi_s$
remains $\forall^2\exists$. A similar
observation holds when $s$ is an $\existsl$ expression nested inside a
top-level {bind} expression.  Since $\existsl$ is generated as a
result of stabilizing a SQL command transformer, which is always a
non-nested bind expression, the subexpression ($s'$) of $\existsl$ is
a non-nested bind expression. $s'$ is however nested inside a
top-level bind expression, hence its prenex signature is
$\forall^2\exists$.  Since $\existsl$ doesn't extend $\nubar$, the
prenex signature of $s$ remains $\forall^2\exists$. When $s$ is an
expression other than $\bind$ or $\existsl$, then $\phi_s$ is not a
quantified formula, and its prenex signature is trivially subsumed by
$\forall^2\exists$. Thus, for the subset of $\SL$, where bind
expressions are restricted to one level of nesting, the FOL formulas
generated by the encoding have the prenex signature as
$\forall^2\exists$.

### 2017-11-06 motivation.tex
Typos etc.

Fig.~\ref{fig:foreach_code} shows a (simplified) snippet of code taken
from Fig.~\ref{fig:new_order_code}. Some irrelevant expressions have
been replaced with constants (\C{k1},\C{k2}, and \C{k3}).  The body of
the loop executes a SQL update followed by an insert.  Recall that a
transaction reads from the global database ($\Delta$), and writes to a
transaction-local database ($\delta$) before committing these
updates. An update statement filters the records that match the search
criteria from $\Delta$ and computes the updated records that are to be
added to the local database. Thus, the state transformer for the
update statement (call it $\T_U$) is the following function on
sets\footnote{$\bind$ has higher precedence than $\cup$; angle braces
($\langle \ldots \rangle$) are used to denote records in math}:
\begin{smathpar}
\begin{array}{l}
  \lambda(\stl,\stg).~ \stl \cup \stg \bind(\lambda \C{s}. 
      \itel{{\sf table}(\C{s}) = \C{Stock} \conj 
        \C{s}.\C{s\_i\_id} = \C{item\_req.ol\_i\_id}\\\hspace*{1.15in}}
           {\{ \langle \C{s\_i\_id}=\C{s.s\_i\_id};\, 
                       \C{s\_d\_id}=\C{s.s\_d\_id};\,
                       \C{s\_qty} = \C{k1}\rangle \}\\\hspace*{1.15in}}
           {\emptyset})
\end{array}
\end{smathpar}
% \begin{smathpar}
% \begin{array}{c}
%   \lambda\Delta.\lambda\delta.\, \delta \;\cup\; 
%       \Delta \,\bind\, (\lambda s.\, 
%           \C{if}\;{\C{stock}(s) \conj 
%                    \C{s.s\_i\_id} = \C{item\_req.ol\_i\_id}}\\
%            \hspace*{0.9in}
%            \C{then}\;{\{\{s \;\C{with}\; \C{s\_qty} = \C{k1}\}\}}\;
%            \C{else}\;{\emptyset})
% \end{array}
% \end{smathpar}
% \begin{verbatim}
%   Rmem(σ'(s')) = Rmem(σ(s')) U 
%                  Rmem[\x. Rmem[\y. if y<x then {y+x} else {}](σ(s)) U {x}](xs)
% \end{verbatim}
% Observe that
% $\T_U(\Delta,\delta)$ is of the form $\delta \,\cup\,
% F_U(\Delta)$.
Here, the set bind operator extracts record elements ($\C{s}$) from
the database, checks the precondition of the update action, and if
satisfied, constructs a new set containing a single record that is
identical to $\C{s}$ except that it binds field $\C{s\_qty}$ to value
$\C{k_1}$.  This new set is added (via set union) to the existing
local database state $\delta$.\footnote{For now, assume that the
  record being added to $\stl$ is not already present in $\stl$.}

@@

\end{array}
\end{smathpar}
Observe that both transformers are of the form $\T(\delta,\Delta) =
\stl \cup \F(\stl,\stg)$, where $\F$ is a function that returns the
set of records added  to the transaction-local database ($\stl$). Let
$\F_U$ and $\F_I$ be the corresponding functions for $\T_U$ and $\T_I$
shown above. The state transformation induced by the loop body in
Fig.~\ref{fig:new_order_code} can be expressed as the following
composition of $\F_U$ and $\F_I$:
\begin{smathpar}
\begin{array}{l}
  \lambda(\stl,\stg).~ \stl \cup \F_U(\stl,\stg) \cup \F_I(\stl,\stg)
\end{array}
\end{smathpar}
The transformer for the loop itself can now be computed to be:
\begin{smathpar}
\begin{array}{l}
  \lambda(\stl,\stg).~ \stl \cup \C{item\_reqs}\bind
      (\lambda\C{item\_req}.~ \F_U(\stl,\stg) \cup \F_I(\stl,\stg))
\end{array}
\end{smathpar}
Observe that the structure of the transformer mirrors the structure of
the program itself. In particular, SQL statements become set
operations, and the \C{foreach} combinator becomes set monad's bind
($\bind$) combinator.  As we demonstrate, the advantage of inferring
such transformers is that we can now make use of a
semantics-preserving translation from the domain of sets equipped with
$\bind$ to a decidable fragment of first-order logic, allowing us to
leverage SMT solvers for automated proofs without having to infer
potentially complex thread-local invariants or intermediate
assertions.  Sec.~\ref{sec:inference} describes this translation. In
the exposition thus far, we assumed $\Delta$ remains invariant, which
is clearly not the case when we admit concurrency.  Necessary
concurrency extensions of the state transformer semantics to deal with
interference is also covered in Sec.~\ref{sec:inference}.  Before
presenting the transformer semantics, we first focus our attention in
the following two sections on the theoretical foundations for weak
isolation, upon which this semantics is based.

>>>>

Fig.~\ref{fig:foreach_code} shows a (simplified) snippet of code taken
from Fig.~\ref{fig:new_order_code}. Some irrelevant expressions have
been replaced with constants (\C{k1},\C{k2}, and \C{k3}).  The body of
the loop executes a SQL update followed by an insert.  Recall that a
transaction reads from the global database ($\Delta$), and writes to a
transaction-local database ($\delta$) before committing these
updates. An update statement filters the records that match the search
criteria from $\Delta$ and computes the updated records that are to be
added to the local database. Thus, the state transformer for the
update statement (call it $\T_U$) is the following function on
sets\footnote{Bind ($\bind$) has higher precedence than union
($\cup$). Angle braces ($\langle \ldots \rangle$) are used to denote
records in math}:
\begin{smathpar}
\begin{array}{l}
  \lambda(\stl,\stg).~ \stl \cup \stg \bind(\lambda \C{s}. 
      \itel{{\sf table}(\C{s}) = \C{Stock} \conj 
        \C{s}.\C{s\_i\_id} = \C{item\_req.ol\_i\_id}\\\hspace*{1.15in}}
           {\{ \langle \C{s\_i\_id}=\C{s.s\_i\_id};\, 
                       \C{s\_d\_id}=\C{s.s\_d\_id};\,
                       \C{s\_qty} = \C{k1}\rangle \}\\\hspace*{1.15in}}
           {\emptyset})
\end{array}
\end{smathpar}
% \begin{smathpar}
% \begin{array}{c}
%   \lambda\Delta.\lambda\delta.\, \delta \;\cup\; 
%       \Delta \,\bind\, (\lambda s.\, 
%           \C{if}\;{\C{stock}(s) \conj 
%                    \C{s.s\_i\_id} = \C{item\_req.ol\_i\_id}}\\
%            \hspace*{0.9in}
%            \C{then}\;{\{\{s \;\C{with}\; \C{s\_qty} = \C{k1}\}\}}\;
%            \C{else}\;{\emptyset})
% \end{array}
% \end{smathpar}
% \begin{verbatim}
%   Rmem(σ'(s')) = Rmem(σ(s')) U 
%                  Rmem[\x. Rmem[\y. if y<x then {y+x} else {}](σ(s)) U {x}](xs)
% \end{verbatim}
% Observe that
% $\T_U(\Delta,\delta)$ is of the form $\delta \,\cup\,
% F_U(\Delta)$.
Here, the set bind operator extracts record elements ($\C{s}$) from
the database, checks the precondition of the update action, and if
satisfied, constructs a new set containing a single record that is
identical to $\C{s}$ except that it binds field $\C{s\_qty}$ to value
$\C{k_1}$.  This new set is added (via set union) to the existing
local database state $\delta$.\footnote{For now, assume that the
  record being added to $\stl$ is not already present in $\stl$.}

@@

\end{array}
\end{smathpar}
Observe that both transformers are of the form $\T(\delta,\Delta) =
\stl \cup \F(\stg)$, where $\F$ is a function that returns the
set of records added  to the transaction-local database ($\stl$). Let
$\F_U$ and $\F_I$ be the corresponding functions for $\T_U$ and $\T_I$
shown above. The state transformation induced by the loop body in
Fig.~\ref{fig:new_order_code} can be expressed as the following
composition of $\F_U$ and $\F_I$:
\begin{smathpar}
\begin{array}{l}
  \lambda(\stl,\stg).~ \stl \cup \F_U(\stg) \cup \F_I(\stg)
\end{array}
\end{smathpar}
The transformer for the loop itself can now be computed to be:
\begin{smathpar}
\begin{array}{l}
  \lambda(\stl,\stg).~ \stl \cup \C{item\_reqs}\bind
      (\lambda\C{item\_req}.~ \F_U(\stg) \cup \F_I(\stg))
\end{array}
\end{smathpar}
Observe that the structure of the transformer mirrors the structure of
the program itself. In particular, SQL statements become set
operations, and the \C{foreach} combinator becomes set monad's bind
($\bind$) combinator.  As we demonstrate, the advantage of inferring
such transformers is that we can now make use of a
semantics-preserving translation from the domain of sets equipped with
$\bind$ to a decidable fragment of first-order logic, allowing us to
leverage SMT solvers for automated proofs without having to infer
potentially complex thread-local invariants or intermediate
assertions.  Sec.~\ref{sec:inference} describes this translation. In
the exposition thus far, we assumed $\Delta$ remains invariant, which
is clearly not the case when we admit concurrency.  Necessary
concurrency extensions of the state transformer semantics to deal with
interference is also covered in Sec.~\ref{sec:inference}.  Before
presenting the transformer semantics, we first focus our attention in
the following two sections on the theoretical foundations for weak
isolation, upon which this semantics is based.
